An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals

### 2021-11-08 analysis/paper/paper.Rmd
Draft v2

---
title: "An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals"
author:
  - Max F. Czapanskiy:
      email: maxczap@stanford.edu
      institute: [Hopkins]
      correspondence: true
  - Paul J. Ponganis:
      institute: [Scripps]
      correspondence: false
  - James A. Fahlbusch:
      institute: [Hopkins]
      correspondence: false
  - T. L. Schmitt:
      institute: [SeaWorld]
      correspondence: false
  - Jeremy A. Goldbogen:
      institute: [Hopkins]
      correspondence: false
institute:
  - Hopkins: Hopkins Marine Station, Department of Biology, Stanford University, USA
  - Scripps: Scripps Institution of Oceanography, University of California San Diego, USA
  - SeaWorld: Veterinary Services, SeaWorld of California, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx"
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: [zotero.bib,references.bib]
csl: "../templates/the-journal-of-experimental-biology.csl"
abstract: |
  Physio-logging methods, which use animal-borne devices to record physiological variables, are entering a new era driven by advances in sensor development. However, existing datasets collected with traditional bio-loggers, such as accelerometers, still contain untapped eco-physiological information. Here we present a computational method for extracting heartrate from high-resolution accelerometer data using a ballistocardiogram. We validated our method with simultaneous accelerometer-electrocardiogram tag deployments in a controlled setting on a killer whale (*Orcinus orca*) and demonstrate the method recovers previously observed cardiovascular patterns in a blue whale (*Balaenoptera musculus*), including the magnitude of apneic bradycardia and progressive relaxation of bradycardia during dives. Our ballistocardiogram method may be applied to mine heart rates from previously collected accelerometery and expand our understanding of comparative cardiovascular physiology.
keywords: |
  ballistocardiogram; diving physiology; cardiovascular physiology; accelerometer; bio-logging; marine mammal
---

@@

Recent advances in physio-logging (recording physiological variables using animal-borne devices) have largely been driven by new developments in sensor technology [@hawkesIntroductionThemeIssue2021]. For example, new physio-logging tags can detect regional changes in blood flow by incorporating functional near-infrared spectroscopy sensors [@mcknightShiningNewLight2021]. However, traditional inertial measurement unit (IMU) tags equipped with accelerometers and other inertial sensors can also measure important physiological and related variables. Through careful inspection and analysis of high-resolution acceleration, scientists have measured elevated respiration rates following record-breaking dives [@satoStrokeRatesDiving2011], near-continuous feeding by small cetaceans [@wisniewskaUltraHighForagingRates2016], social interactions between large cetaceans  [@goldbogenUsingAccelerometersDetermine2014], and important biomechanical variables including movement speed [@cadeDeterminingForwardSpeed2018]. While physio-logging tags with cutting-edge biomedical technologies push the boundaries of physiological field research, simpler IMU tags have fewer logistical constraints and provide access to more species and larger sample sizes. This is particularly important for species that cannot be restrained or studied in managed care. For example, of the sixteen species of baleen whales (Mysticeti), heart rate has only been recorded with an electrocardiogram tag in the wild for one blue whale (_Balaenoptera musculus_) [@goldbogenExtremeBradycardiaTachycardia2019; but see @ponganisHeartRateElectrocardiogram1999]. Conversely, IMU tags have been deployed on hundreds of individuals of nearly every species in the clade for the last twenty years [@nowacekBuoyantBalaenidsUps2001]. These existing datasets (and future IMU tag deployments) could hold additional valuable physiological information, awaiting proper computational methods for mining them.

The ballistocardiogram (BCG) has potential applications to using accelerometers as heartrate monitors in both the wild and in managed care. Ballistocardiography is a noninvasive method for measuring cardiac function based on the ballistic forces involved in the heart ejecting blood into the major vessels. The BCG originated as a clinical tool in the first half of the 20th century [@starrStudiesEstimationCardiac1939], but was largely superseded by electro- and echocardiography. However the medical community has recently returned to ballistocardiography as a potential means of passive monitoring of heart function in at-risk populations [@giovangrandi2011ballistocardiography], which has led to substantial progress in signal processing methodology for generating interpretable BCGs [@sadekBallistocardiogramSignalProcessing2019]. While the BCG is a three-dimensional phenomenon, it is strongest in the anterior-posterior axis [@inanBallistocardiographySeismocardiographyReview2015]. Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these are the IJK complex, which progressively occurs during systole [@pinheiroTheoryDevelopmentsUnobtrusive2010]. The BCG J wave is the most robust feature in the waveform and typically used for detecting heart beats.

Here we present a method for generating a BCG from bio-logger accelerometry. We validated our method with a simultaneously recorded electrocardiogram (ECG) on an adult killer whale in managed care (_Orcinus orca_) and applied it to detect heartrate in a blue whale. The relative orientation of the tag on the body is uncertain in cetacean bio-logging in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on anterior-posterior acceleration (surge), we also generated a three-dimensional BCG, which we expected would be more robust in a field setting. Specifically, we tested three hypotheses to validate our method. First, a one-dimensional BCG would, in a controlled setting, produce statistically equivalent instantaneous heartrates as an ECG. Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG. Third, BCG-derived heartrates would increase during the bottom and latter phases of dives, consistent with the progressive increase in heartrate routinely observed before and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].

@@

A killer whale in managed care at the SeaWorld, San Diego, CA marine facilities was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld display permit. The CATS tag recorded acceleration at 400 Hz, magnetometer and gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps. All sensors were rotated from the tag's frame of reference to that of the whale using MATLAB (MathWorks, Inc., v2020b) tools for processing CATS data [@cadeToolsIntegratingInertial2021]. The ECG tag hardware and data processing followed the methods in [@bickettHeartRatesHeart2019]. Briefly, we recorded ECG at 100 Hz and identified individual heart beats from visually verified R-waves using a customized peak detection program (K. Ponganis; Origin 2017, OriginLab Co., Northampton, MA). ECG and IMU were recorded during a submerged breath hold.

@@

The BCG waveform is three dimensional, but strongest in the anterior-posterior axis [@inanBallistocardiographySeismocardiographyReview2015]. We tested 1d and 3d metrics for identifying heartbeats in acceleration data based on the methods of [@leePhysiologicalSignalMonitoring2016]. For windowed operations, we used 0.5 s for killer whale data and 2.0 s for blue whale data.

@@

1. Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) [R package `signal` @R-signal].
2. Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter (R package `signal`). Differentiation exaggerates impulses like the J wave.
3. Further enhance the peaks by calculating the Shannon entropy ($H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$, where $k$ is the acceleration axis). Additionally, the Shannon entropy is strictly positive, which facilitates peak detection. In the 1d case, $k$ is surge (anterior-posterior acceleration) only.
4. Remove noise by applying a triangular moving average smoother.
5. Extract peaks and heuristically remove noisy peaks (Fig. S1).

This procedure can be applied to either 1d (i.e., surge-only) or 3d acceleration. In the case of 3d acceleration, the band-pass and Savitzky-Golay filters were applied to each axis independently.

## ECG validation on killer whale

We fit ordinary least squares regression to BCG-derived instantaneous heart rates with respect to ECG-derived and tested 1) if the intercept was significantly different than 0 and 2) if the slope was significantly different than 1. We calculated the mean and standard deviation of absolute error as an equivalence measure (1d BCG only).

@@

Dynamic body movements produce an acceleration signal that masks the ballistocardiogram, so we limited our analyses to motionless periods (Fig. S2). These periods occured during or near the bottom phase of dives between fluke strokes. Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [_sensu_ @goughScalingSwimmingPerformance2019].

We tested whether the 3d BCG was more robust than 1d BCG in field data by comparing the signal-to-noise ratios. For both BCGs, we calculated the power spectral density [R package `psd`; @Barbour2014]. Previously recorded blue whale apneic heart rate was 4-8 bpm [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified _signal_ as the integration of the power spectral density curve from 4-8 bpm and _noise_ as the integrated remainder, up to 60 bpm.

We also tested whether BCG-derived instantaneous heart rates exhibited bradycardia relaxation over the course of dives, consistent with diving physiology patterns in marine mammals [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014]. We assigned dive start and end times when the whale swam deeper than 2 m, retaining dives that exceeded 10 m depth and 5 minutes duration. Dive times were normalized from 0 (start of dive) to 1 (end of dive). We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (to account for heteroscedascity) [R package `RobustLinearReg`; @R-RobustLinearReg; @Sen-1968; @theilRankInvariantMethodLinear1992] and tested whether the slope was greater than 0.

@@

## ECG validation on killer whale

@@

The ECG and BCG yielded nearly identical heart rate estimations (Fig. \@ref(fig:oo-bcg-ecg)). We collected 14 s of simultaneous ECG and BCG data during a motionless, submerged breath hold. BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean &#177; standard deviation). Ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean &#177; standard error), which were not significantly different from the hypothesized 1 and 0, respectively.

@@

We generated 1d and 3d BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig. S3).

@@

The 3d BCG (Fig. \@ref(fig:bw-bcg)) produced a more robust signal than the 1d BCG, which used only anterior-posterior acceleration. The signal-to-noise ratio was 2.00 for the 3d BCG, compared to 0.17 for the 1d BCG (Fig. \@ref(fig:bw-validation-plots)**A**).

@@

3d BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives. Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig. \@ref(fig:bw-validation-plots)**B**).

@@

* Everyone who helped collect and process the blue whale data.
* The Sea World trainers.
* Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.
* Funding from Office of Naval Research N000141912455, Stanford Terman Fellowship
* M. F. Czapanskiy was supported by the Stanford University William R. and Sara Hart Kimball Fellowship and a Stanford Data Science Scholar Fellowship.

@@

```{r oo-bcg-ecg, fig.cap="The ECG (**A**) and BCG (**E**), recorded by the ECG and CATS IMU tags, respectively, produced nearly identical heart beat predictions for the killer whale. **B-D** display the intermediate steps in the BCG signal processing procedure. **B**: Anterior-posterior axis acceleration (surge) after band-pass filtering. Inset shows the IJK complex with surrounding H and L waves for the region bounded by the dashed box. **C**: Peaks enhanced after forward differencing. **D**: A strictly positive signal after calculating Shannon entropy. Y-axis labeling follows [@leePhysiologicalSignalMonitoring2016]; y-axis values excluded because filtering introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
t <- theme_classic() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_blank())
t_yvoid <- t +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

@@

inset_range <- as.POSIXct("2021-08-16 10:09:29.9", tz = "US/Pacific") + c(0, 4)
pB <- ggplot(oo_400hz, aes(dt, surge_filt)) +
  geom_line(size = 0.2) +
  annotation_custom(grob = ggplotGrob(pB_inset),
                    xmin = inset_range[1], xmax = inset_range[2],
                    ymin = 0.25, ymax = 0.5) +
  annotate("rect",
           xmin = ijk_range[1], 
           xmax = ijk_range[2],
           ymin = min(ijk_complex$surge_filt) - 0.02, 
           ymax = max(ijk_complex$surge_filt) + 0.02,
           fill = NA,
           color = "grey50",
           lty = 3) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  expand_limits(y = 0.5) +
  labs(y = bquote(f[HR])) +
  t_yvoid

# C: Surge diff
pC <- ggplot(oo_400hz, aes(dt, surge_diff)) +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "dBCG") +
  t_yvoid

# D: Shannon entropy
pD <- ggplot(oo_400hz, aes(dt, surge_se)) +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "seBCG") +
  t_yvoid

# E: Smoothed
pE <- ggplot(oo_400hz, aes(dt, surge_smooth)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(oo_400hz, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "Y") +
  t_yvoid

@@

```{r bw-bcg, fig.cap="Example of signal processing for 3d BCG during a motionless period in a blue whale dive. **A**: Band-pass filtered triaxial acceleration, where surge is along the anterior-posterior axis, sway is along the lateral axis, and heave is along the dorso-ventral axis. **B**: Differencing the filtered acceleration enhances peaks. **C**: Calculating the Shannon entropy combines information from all three axes and makes the signal strictly positive. **D**: Smoothing the Shannon entropy facilitates robust peak detection. Detected heart beats in blue. Y-axis labeling follows [@leePhysiologicalSignalMonitoring2016]; y-axis values were excluded because the filtering process introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
bw_bcg_plot <- function(rid) {
  bcg_sample <- filter(bw_400hz, region_id == rid)

  # A: Filtered surge
  pA <- bcg_sample %>% 
    rename(Surge = surge_filt, Sway = sway_filt, Heave = heave_filt) %>% 
    pivot_longer(Surge:Heave, names_to = "axis", values_to = "acc") %>% 
    mutate(axis = factor(axis, levels = c("Surge", "Sway", "Heave"))) %>% 
    ggplot(aes(dt, acc)) +
    geom_line(size = 0.2) +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_y_continuous(n.breaks = 3) +
    facet_grid(rows = vars(axis)) +
    labs(y = bquote(f[HR])) +
    t_yvoid

  # B: Jerk vectors
  pB <- bcg_sample %>% 
    mutate(jerk = set_names(as_tibble(jerk), c("Surge", "Sway", "Heave"))) %>% 
    unpack(jerk) %>% 
    pivot_longer(Surge:Heave, names_to = "axis", values_to = "jerk") %>% 
    mutate(axis = factor(axis, levels = c("Surge", "Sway", "Heave"))) %>% 
    ggplot(aes(dt, jerk)) +
    geom_line(size = 0.2) +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_y_continuous(n.breaks = 3) +
    facet_grid(rows = vars(axis)) +
    labs(y = "dBCG") +
    t_yvoid

  # C: Shannon entropy of jerk (NOT jerk norm)
  pC <- ggplot(bcg_sample, aes(dt, jerk_se)) +
    geom_line(size = 0.2) +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = "seBCG") +
    t_yvoid

  # D: Smoothed
  pD <- ggplot(bcg_sample, aes(dt, jerk_smooth)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = "Y") +
    t_yvoid

@@

<div id="refs"></div>

>>>>

---
title: "An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals"
author:
  - Max F. Czapanskiy:
      email: maxczap@stanford.edu
      institute: [Hopkins]
      correspondence: true
  - Paul J. Ponganis:
      institute: [Scripps]
      correspondence: false
  - James A. Fahlbusch:
      institute: [Hopkins]
      correspondence: false
  - T. L. Schmitt:
      institute: [SeaWorld]
      correspondence: false
  - Jeremy A. Goldbogen:
      institute: [Hopkins]
      correspondence: false
institute:
  - Hopkins: Hopkins Marine Station, Department of Biology, Stanford University, USA
  - Scripps: Scripps Institution of Oceanography, University of California San Diego, USA
  - SeaWorld: Veterinary Services, SeaWorld of California, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx"
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: [zotero.bib,references.bib]
csl: "../templates/the-journal-of-experimental-biology.csl"
abstract: |
  Physio-logging methods, which use animal-borne devices to record physiological variables, are entering a new era driven by advances in sensor development. However, existing datasets collected with traditional bio-loggers, such as accelerometers, still contain untapped eco-physiological information. Here we present a computational method for extracting heartrate from high-resolution accelerometer data using a ballistocardiogram. We validated our method with simultaneous accelerometer-electrocardiogram tag deployments in a controlled setting on a killer whale (*Orcinus orca*) and demonstrate the method recovers previously observed cardiovascular patterns in a blue whale (*Balaenoptera musculus*), including the magnitude of apneic bradycardia and increase in heart rate prior to and during ascent. Our ballistocardiogram method may be applied to mine heart rates from previously collected accelerometery and expand our understanding of comparative cardiovascular physiology.
keywords: |
  ballistocardiogram; diving physiology; cardiovascular physiology; accelerometer; bio-logging; marine mammal
---

@@

Recent advances in physio-logging (recording physiological variables using animal-borne devices) have largely been driven by new developments in sensor technology [@hawkesIntroductionThemeIssue2021]. For example, new physio-logging tags can detect regional changes in blood flow by incorporating functional near-infrared spectroscopy sensors [@mcknightShiningNewLight2021]. However, traditional inertial measurement unit (IMU) tags equipped with accelerometers and other inertial sensors can also measure important physiological and related variables. Through careful inspection and analysis of high-resolution acceleration, scientists have measured elevated respiration rates following record-breaking dives [@satoStrokeRatesDiving2011], near-continuous feeding by small cetaceans [@wisniewskaUltraHighForagingRates2016], social interactions between large cetaceans [@goldbogenUsingAccelerometersDetermine2014], and important biomechanical variables including movement speed [@cadeDeterminingForwardSpeed2018]. While physio-logging tags with cutting-edge biomedical technologies push the boundaries of physiological field research, simpler IMU tags have fewer logistical constraints and provide access to more species and larger sample sizes. This is particularly important for species that cannot be restrained or studied in managed care. For example, of the sixteen species of baleen whales (Mysticeti), heart rate has only been recorded with an electrocardiogram tag in the wild for one blue whale (*Balaenoptera musculus*) [@goldbogenExtremeBradycardiaTachycardia2019; but see @ponganisHeartRateElectrocardiogram1999]. Conversely, IMU tags have been deployed on hundreds of individuals of nearly every species in the clade for the last twenty years [@nowacekBuoyantBalaenidsUps2001]. These existing datasets (and future IMU tag deployments) could hold additional valuable physiological information, awaiting proper computational methods for mining them.

The ballistocardiogram (BCG) has potential applications to using accelerometers as heartrate monitors in both the wild and in managed care. Ballistocardiography is a noninvasive method for measuring cardiac function based on the ballistic forces involved in the heart ejecting blood into the major vessels. The BCG originated as a clinical tool in the first half of the 20th century [@starrStudiesEstimationCardiac1939], but was largely superseded by electro- and echocardiography. However the medical community has recently returned to ballistocardiography as a potential means of passive monitoring of heart function in at-risk populations [@giovangrandi2011ballistocardiography], which has led to substantial progress in signal processing methodology for generating interpretable BCGs [@sadekBallistocardiogramSignalProcessing2019]. While the BCG is a three-dimensional phenomenon, it is strongest in the cranio-caudal axis [@inanBallistocardiographySeismocardiographyReview2015]. Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these are the IJK complex, which progressively occurs during systole [@pinheiroTheoryDevelopmentsUnobtrusive2010]. The BCG J wave is the most robust feature in the waveform and typically used for detecting heart beats.

Here we present a method for generating a BCG from bio-logger accelerometry. We validated our method with a simultaneously recorded electrocardiogram (ECG) on an adult killer whale in managed care (*Orcinus orca*) and applied it to detect heartrate in a blue whale. The relative orientation of the tag on the body is uncertain in cetacean bio-logging in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on cranio-caudal acceleration, we also generated a three-dimensional BCG, which we expected would be more robust in a field setting. Specifically, we tested three hypotheses to validate our method. First, a one-dimensional BCG would, in a controlled setting, produce statistically equivalent instantaneous heartrates as an ECG. Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG. Third, BCG-derived heartrates would increase during the latter phases of dives, consistent with the progressive increase in heartrate routinely observed prior to and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].

@@

A 3807 kg (**needs confirmation**) killer whale in managed care at the SeaWorld, San Diego, CA marine facilities was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld display permit. The CATS tag recorded acceleration at 400 Hz, magnetometer and gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps. All sensors were rotated from the tag's frame of reference to that of the whale using MATLAB (MathWorks, Inc., v2020b) tools for processing CATS data [@cadeToolsIntegratingInertial2021]. This rotation aligned the tag's x-, y-, and z- axes with the cranio-caudal, lateral, and dorso-ventral axes of the whale, respectively. The ECG tag hardware and data processing followed the methods in [@bickettHeartRatesHeart2019]. Briefly, we attached the tag on the mid-lateral left chest posterior to the pectoral fin and recorded ECG at 100 Hz. Individual heart beats were identified from visually verified R-waves using a customized peak detection program (K. Ponganis; Origin 2017, OriginLab Co., Northampton, MA). ECG and IMU were recorded during a spontaneous breath hold while the whale rested at the surface.

@@

The BCG waveform is three dimensional, but strongest in the cranio-caudal axis [@inanBallistocardiographySeismocardiographyReview2015]. We tested both 1-dimensional (cranio-caudal only) and 3-dimensional metrics for identifying heartbeats in acceleration data based on the methods of [@leePhysiologicalSignalMonitoring2016]. For windowed operations, we used 0.5 s for killer whale data and 2.0 s for blue whale data.

@@

1.  Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) [R package `signal` @R-signal].
2.  Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter (R package `signal`). Differentiation exaggerates impulses like the J wave.
3.  Further enhance the peaks by calculating the Shannon entropy ($H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$, where $k$ is the acceleration axis). Additionally, the Shannon entropy is strictly positive, which facilitates peak detection. In the 1-dimensional case, $k$ is surge (cranio-caudal acceleration) only.
4.  Remove noise by applying a triangular moving average smoother.
5.  Extract peaks and heuristically remove noisy peaks (Fig. S1).

This procedure may be applied to either 1-dimensional (i.e., cranio-caudal only) or 3-dimensional acceleration. In the case of 3-dimensional acceleration, the band-pass and Savitzky-Golay filters were applied to each axis independently.

## BCG validation with killer whale ECG

We fit ordinary least squares regression to BCG-derived instantaneous heart rates with respect to ECG-derived and tested 1) if the intercept was significantly different than 0 and 2) if the slope was significantly different than 1. We calculated the mean and standard deviation of absolute error as an equivalence measure (1-dimensional BCG only).

@@

Dynamic body movements produce an acceleration signal that masks the ballistocardiogram, so we limited our analyses to motionless periods (Fig. S2). These periods occured during or near the bottom phase of dives between fluke strokes. Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [*sensu* @goughScalingSwimmingPerformance2019].

We tested whether the 3-dimensional BCG was more robust than 1-dimensional BCG in field data by comparing the signal-to-noise ratios. For both BCGs, we calculated the power spectral density [R package `psd`; @Barbour2014]. Previously recorded blue whale apneic heart rate was 4-8 beats per minute (bpm) [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified *signal* as the integration of the power spectral density curve from 4-8 bpm and *noise* as the integrated remainder, up to 60 bpm.

We also tested whether BCG-derived instantaneous heart rates were consistent with the range and pattern of heart rates previously observed in the blue whale and other marine mammals; namely a gradual increase in heart rate later in the dive, especially during the final ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014]. We assigned dive start and end times when the whale swam deeper than 2 m, retaining dives that exceeded 10 m depth and 5 minutes duration. Dive times were normalized from 0 (start of dive) to 1 (end of dive). We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (to account for heteroscedascity) [R package `RobustLinearReg`; @R-RobustLinearReg; @Sen-1968; @theilRankInvariantMethodLinear1992] and tested whether the slope was greater than 0.

@@

## BCG validation with killer whale ECG

@@

The ECG and BCG yielded nearly identical heart rate estimations (Fig. \@ref(fig:oo-bcg-ecg)). We collected 14 s of simultaneous ECG and BCG data during a motionless, submerged breath hold. BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean ± standard deviation). Ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean ± standard error), which were not significantly different from the hypothesized 1 and 0, respectively.

@@

We generated 1-dimensional and 3-dimensional BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig. S3).

@@

The 3-dimensional BCG (Fig. \@ref(fig:bw-bcg)) produced a more robust signal than the 1-dimensional BCG, which used only cranio-caudal acceleration. The signal-to-noise ratio was 2.00 for the 3-dimensional BCG, compared to 0.17 for the 1-dimensional BCG (Fig. \@ref(fig:bw-validation-plots)**A**).

@@

3-dimensional BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives. Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig. \@ref(fig:bw-validation-plots)**B**).

@@

-   Everyone who helped collect and process the blue whale data.
-   The Sea World trainers.
-   Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.
-   Funding from Office of Naval Research N000141912455, Stanford Terman Fellowship
-   M. F. Czapanskiy was supported by the Stanford University William R. and Sara Hart Kimball Fellowship and a Stanford Data Science Scholar Fellowship.

@@

```{r oo-bcg-ecg, fig.cap="The ECG (**A**, recorded by ECG tag) and 1-dimensional BCG (**E**, processed from the cranio-caudal acceleration recorded by the IMU tag) produced nearly identical heart beat predictions for the killer whale. **B-D** display the intermediate steps in the BCG signal processing procedure. **B**: Cranio-caudal axis acceleration after band-pass filtering. Inset shows the IJK complex with surrounding H and L waves for the region bounded by the dashed box. **C**: Peaks enhanced after forward differencing acceleration (i.e., jerk). **D**: A strictly positive signal after calculating Shannon entropy. Y-axis values excluded because filtering introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
t <- theme_classic() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_blank())
t_yvoid <- t +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

@@

inset_range <- as.POSIXct("2021-08-16 10:09:29.9", tz = "US/Pacific") + c(0, 4)
pB <- ggplot(oo_400hz, aes(dt, surge_filt)) +
  geom_line(size = 0.2) +
  annotation_custom(grob = ggplotGrob(pB_inset),
                    xmin = inset_range[1], xmax = inset_range[2],
                    ymin = 0.25, ymax = 0.5) +
  annotate("rect",
           xmin = ijk_range[1], 
           xmax = ijk_range[2],
           ymin = min(ijk_complex$surge_filt) - 0.02, 
           ymax = max(ijk_complex$surge_filt) + 0.02,
           fill = NA,
           color = "grey50",
           lty = 3) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  expand_limits(y = 0.5) +
  labs(y = "Cranio-caudal acceleration (filtered)") +
  t_yvoid

# C: Surge diff
pC <- ggplot(oo_400hz, aes(dt, surge_diff)) +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "Cranio-caudal jerk") +
  t_yvoid

# D: Shannon entropy
pD <- ggplot(oo_400hz, aes(dt, surge_se)) +
  geom_line(size = 0.2) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "Shannon entropy") +
  t_yvoid

# E: Smoothed
pE <- ggplot(oo_400hz, aes(dt, surge_smooth)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(oo_400hz, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "BCG") +
  t_yvoid

@@

```{r bw-bcg, fig.cap="Example of signal processing for 3-dimensional BCG during a motionless period in a blue whale dive. **A**: Band-pass filtered triaxial acceleration, where surge is along the cranio-caudal axis, sway is along the lateral axis, and heave is along the dorso-ventral axis. **B**: Differencing the filtered acceleration enhances peaks. **C**: Calculating the Shannon entropy combines information from all three axes and makes the signal strictly positive. **D**: Smoothing the Shannon entropy facilitates robust peak detection. Detected heart beats in blue. Y-axis labeling follows [@leePhysiologicalSignalMonitoring2016]; y-axis values were excluded because the filtering process introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
bw_bcg_plot <- function(rid) {
  bcg_sample <- filter(bw_400hz, region_id == rid)

  # A: Filtered surge
  pA <- bcg_sample %>% 
    rename(`Cranio-caudal` = surge_filt, 
           Lateral = sway_filt, 
           `Dorso-ventral` = heave_filt) %>% 
    pivot_longer(`Cranio-caudal`:`Dorso-ventral`, 
                 names_to = "axis", 
                 values_to = "acc") %>% 
    mutate(axis = factor(axis, levels = c("Cranio-caudal", 
                                          "Lateral", 
                                          "Dorso-ventral"))) %>% 
    ggplot(aes(dt, acc)) +
    geom_line(size = 0.2) +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_y_continuous(n.breaks = 3) +
    facet_grid(rows = vars(axis)) +
    labs(y = "Acceleration (filtered)") +
    t_yvoid

  # B: Jerk vectors
  pB <- bcg_sample %>% 
    mutate(jerk = set_names(as_tibble(jerk), c("Cranio-caudal", 
                                               "Lateral", 
                                               "Dorso-ventral"))) %>% 
    unpack(jerk) %>% 
    pivot_longer(`Cranio-caudal`:`Dorso-ventral`, 
                 names_to = "axis", 
                 values_to = "jerk") %>% 
    mutate(axis = factor(axis, levels = c("Cranio-caudal", 
                                          "Lateral", 
                                          "Dorso-ventral"))) %>% 
    ggplot(aes(dt, jerk)) +
    geom_line(size = 0.2) +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    scale_y_continuous(n.breaks = 3) +
    facet_grid(rows = vars(axis)) +
    labs(y = "Jerk") +
    t_yvoid

  # C: Shannon entropy of jerk (NOT jerk norm)
  pC <- ggplot(bcg_sample, aes(dt, jerk_se)) +
    geom_line(size = 0.2) +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = "Shannon entropy") +
    t_yvoid

  # D: Smoothed
  pD <- ggplot(bcg_sample, aes(dt, jerk_smooth)) +
    geom_line(size = 0.2) +
    geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
    scale_x_datetime(date_labels = "%H:%M:%S") +
    labs(y = "BCG") +
    t_yvoid

@@

::: {#refs}
:::
